---
title: "MA615-EDA-Strawberry"
author: "Mingrui Du"
format: pdf
editor: visual
---

## Overview

### Objective

The objective of this assignment is to practice data cleaning then exploratory data analysis(EDA) on data set "Strawberry"

```{r}
#| echo: false
suppressMessages(library(dplyr))
library(knitr)  
#library(kableExtra)
library(tidyverse)
library(stringr)
```

### Data acquisition

"Strawberry" selected from [USDA NASS](https://quickstats.nass.usda.gov/) , original data stored [here](https://quickstats.nass.usda.gov/results/45FBC825-B104-38E2-9802-839F5F3C7036)

### The Data

The original data set consists of 21 columns, among which "Week.Ending","Ag.District", "Ag.District.Code","County","County.ANSI","Zip.Code","Region","Watershed" have only NA elements, and "Geo.Level"(=STATE), "watershed_code"(=0),"Commodity"(=STRAWBERRIES) are single-value columns.

Column "Program" contains two sources of data: CENSUS and SURVEY.

"Data.Item" is a concatenation of six mixed columns:

-   Commodity = STRAWBERRIES in our case

-   class_desc: recording a physical attribute (not recorded in this strawberry case)

-   prodn_practice_desc: a method of production or action taken on the commodity(e.g., IRRIGATED, ORGANIC, ON FEED)

-   util_practice_desc: Utilizations (e.g., GRAIN, FROZEN, SLAUGHTER) or marketing channels (e.g., FRESH MARKET, PROCESSING, RETAIL)

-   statisticcat_desc: The aspect of a commodity being measured (e.g., AREA HARVESTED, PRICE RECEIVED, INVENTORY, SALES).

-   unit.

"Domain":

-   domain = ORGANIC STATUS for organic commodity;

-   for chemical usage data, the domain describes the type of chemical applied to the commodity(e.g., FUNGICIDE, HERBICIDE, INSECTICIDE, FERTILIZER, OTHER);

-   domain = TOTAL will have no further breakouts.

"Domain.Category" records categories or partitions within a domain. For instance, the specific chemical taken for a commodity, along with PC code.

"Value" contains published data value or [suppression reason code](https://quickstats.nass.usda.gov/src/glossary.pdf).

### Data assessment

"Data.Item" compresses multiple columns into one; "Domain.Category" contains various chemicals(and PC codes) that require extraction; "Value" involves numbers in different units, as well as abbreviation codes, so does "CV".

Here I first delete single value columns then split dataset into CENSUS and SURVEY, since organic commodity lies in CENSUS and chemical usage stored in the latter.  

### Initial questions

-   Initial questions about strawberries, the data, and about the work you are undertaking. Write these before you begin working.

## Data cleaning and organization

<!-- Remove single-value columns -->

```{r}
#| echo: false
strawberry <- read.csv("strawberry.csv", header = TRUE)
strwb <- strawberry ## copy of raw data
## drop single value columns
strawberry <- strawberry %>% 
              select_if(~!all(is.na(.))) %>% 
              select(-Geo.Level, -watershed_code, -Commodity)
```

<!-- Separate CENSUS and SURVEY -->

```{r}
## split CENSUS and SURVEY
strwb_census <- strawberry |> filter(Program == "CENSUS")
strwb_survey <- strawberry |> filter(Program == "SURVEY")
#nrow(strawberry) == (nrow(strwb_census) + nrow(strwb_survey))
```

<!-- Start with CENSUS -->

<!-- Clean composite column: Data.Item -->

<!-- Column separators: , and - -->

```{r}
## No need to deal with Domain.Category in CENSUS as it's single value.
#unique(strwb_census$Domain.Category)
#unique(strwb_census$Domain)

## Seperate Data.Item into 4 columns at most, the first set to be commodity(STRAWBERRIES)
#unique(strwb_census$Data.Item)
strwb_census <- strwb_census |>
                separate_wider_delim(cols = `Data.Item`,
                                     delim = ",",
                                     names = c("Commodity", ## STRAWBERRIES
                                               "temp1",     
                                               "temp2",
                                               "temp3"),
                                     too_many = "error",
                                     too_few = "align_start")
strwb_census$temp1 <- trimws(strwb_census$temp1)
strwb_census$temp2 <- trimws(strwb_census$temp2)
strwb_census$temp3 <- trimws(strwb_census$temp3)

## Separate temp1 into Production Practice(prodn_practice_desc) and prop_acct
#unique(strwb_census$temp1)
#strwb_census |> distinct(temp1)
strwb_census <- strwb_census |>
  separate_wider_delim(  cols = temp1,
                         delim = " - ",
                         names = c("prodn_practice_desc",
                                   "prop_acct"),
                         too_many = "error",
                         too_few = "align_start"
                       )
#unique(strwb_census$prodn_practice_desc)
#unique(strwb_census$prop_acct)
### We can see now NA appears while cleaning data.

## Separate temp2 into marketing channels(Fresh Market,Processing) and units(unit_desc)
strwb_census <- strwb_census |> mutate(`Fresh Market` = temp2, .after = temp2)
strwb_census$`Fresh Market` <- strwb_census$`Fresh Market` |> 
                                str_replace( "^MEA.*", "")|> 
                                str_replace( "^P.*", "")
strwb_census$`Fresh Market`[is.na(strwb_census$`Fresh Market`)] <- ""
strwb_census$temp2 <- strwb_census$temp2 |> str_replace("^F.*", "")
strwb_census$`Fresh Market` <- strwb_census$`Fresh Market` |> str_replace("^FRESH MARKET - ", "")
#unique(strwb_census$temp2)
strwb_census <- strwb_census |> mutate(`Processing` = temp2, .after = temp2)
strwb_census$`Processing` <- strwb_census$`Processing` |> 
                                str_replace( "^MEA.*", "")
strwb_census$`Processing`[is.na(strwb_census$`Processing`)] <- ""
strwb_census$temp2 <- strwb_census$temp2 |> str_replace("^P.*", "")
strwb_census$`Processing` <- strwb_census$`Processing` |> str_replace("^PROCESSING - ", "")
#unique(strwb_census$temp2)
#unique(strwb_census$Processing)
strwb_census$prop_acct[is.na(strwb_census$prop_acct)] <- "" 
strwb_census$temp2[is.na(strwb_census$temp2)] <- "" 
strwb_census$temp3[is.na(strwb_census$temp3)] <- "" 

## Now combine temp2 and temp3 into a column for units
strwb_census <- strwb_census |> unite(temp2, temp3, col="unit_desc", sep="")
#unique(strwb_census$unit_desc)
strwb_census$unit_desc <- strwb_census$unit_desc |> str_replace("^MEASURED IN ", "")
## These separated column have no NA inside
#unique(!is.na(strwb_census$unit_desc))
#unique(!is.na(strwb_census$prop_acct))
#unique(!is.na(strwb_census$`Fresh Market`))
#unique(!is.na(strwb_census$Processing))

## Remove single-value columns: 
# Program(CENSUS), Commodity(STRAWBERRIES), prodn_practice_desc(ORGANIC), Domain(ORGANIC STATUS), Domain.Category(ORGANIC STATUS: (NOP USDA CERTIFIED))
strwb_census <- strwb_census %>% select_if(~length(unique(.)) > 1)

```

<!-- Format Value column: remove comma in numbers and reserve footnotes-->

```{r}
## function to remove comma in numbers
dcomma <- function(c){
  suppressWarnings({
  xnew = as.numeric(gsub(",", "", c))
  fns = unique(c[is.na(xnew)])
  vtran = list("new_vec" = xnew, "footnotes" = fns)
  return(vtran)
  })
}
c <- data.frame(dcomma(strwb_census$Value))
strwb_census$Value <- c[,1]
strwb_census$fn <- c[,2]

## Separate values by units($, CWT)
strwb_census$Value_USD <- ifelse(strwb_census$unit_desc == "$", strwb_census$Value, "")
strwb_census$Value_CWT <- ifelse(strwb_census$unit_desc == "CWT", strwb_census$Value, "")

# data_USD <- strwb_census |>
#   select(c(Year, State, `Fresh Market`, unit_desc, Value)) |>
#   filter((unit_desc == '$') & (`Fresh Market` == 'SALES') & (Value != "(D)"))

```

<!-- SURVEY -->
<!-- Data.Item -->
```{r}
strwb_survey <- strwb_survey |> 
  separate_wider_delim(  cols = `Data.Item`,
                         delim = ",",
                         names = c("temp1",
                                 "temp2",
                                 "temp3",
                                 "temp4"),
                         too_many = "error",
                         too_few = "align_start"
                       )

strwb_survey <- strwb_survey |>
  separate_wider_delim(  cols = temp1,
                         delim = " - ",
                         names = c("Commodity",
                                 "temp1b"),
                         too_many = "error",
                         too_few = "align_start"
                       )
### delete head & tail spaces
strwb_survey$temp4 <- trimws(strwb_survey$temp4)
strwb_survey$temp2 <- trimws(strwb_survey$temp2)
strwb_survey$temp3 <- trimws(strwb_survey$temp3)

## Divide Data.Item into "Fresh Market"  "Processing" "Bearing" "Utilized"
### Fresh Market
strwb_survey <- strwb_survey |> mutate(`Fresh Market` = temp2, .after = temp2)
strwb_survey$`Fresh Market` <- strwb_survey$`Fresh Market` |> 
                                str_replace( "^MEA.*", "") |> 
                                str_replace( "^P.*", "") |>
                                str_replace( "^N.*", "") |>
                                str_replace( "^U.*", "") |>
                                str_replace( "^B.*", "") 
  
strwb_survey$`Fresh Market`[is.na(strwb_survey$`Fresh Market`)] <- ""
strwb_survey$temp2 <- strwb_survey$temp2 |> str_replace("^F.*", "")
strwb_survey$`Fresh Market` <- strwb_survey$`Fresh Market` |> str_replace("^FRESH MARKET - ", "")
### Processing
strwb_survey <- strwb_survey |> mutate(Processing = temp2, .after = temp2)
strwb_survey$Processing <- strwb_survey$Processing |> 
                                str_replace( "^MEA.*", "") |> 
                                str_replace( "^N.*", "") |>
                                str_replace( "^U.*", "") |>
                                str_replace( "^B.*", "")
strwb_survey$Processing[is.na(strwb_survey$Processing)] <- ""
strwb_survey$temp2 <- strwb_survey$temp2 |> str_replace("^P.*", "")
strwb_survey$Processing <- strwb_survey$Processing |> str_replace("^PROCESSING - ", "")
### UTILIZED
strwb_survey <- strwb_survey |> mutate(Utilized = temp2, .after = temp2)
strwb_survey$Utilized <- strwb_survey$Utilized |> 
                                str_replace( "^MEA.*", "") |> 
                                str_replace( "^N.*", "") |>
                                str_replace( "^B.*", "")
strwb_survey$Utilized[is.na(strwb_survey$Utilized)] <- ""
strwb_survey$temp2 <- strwb_survey$temp2 |> str_replace("^U.*", "")
strwb_survey$Utilized <- strwb_survey$Utilized |> str_replace("^UTILIZED - ", "")
### NOT SOLD
strwb_survey <- strwb_survey |> mutate(`Not Sold` = temp2, .after = temp2)
strwb_survey$`Not Sold` <- strwb_survey$`Not Sold` |> 
                                str_replace( "^MEA.*", "") |> 
                                str_replace( "^B.*", "")
strwb_survey$`Not Sold`[is.na(strwb_survey$`Not Sold`)] <- ""
strwb_survey$temp2 <- strwb_survey$temp2 |> str_replace("^N.*", "")
strwb_survey$`Not Sold` <- strwb_survey$`Not Sold` |> str_replace("^NOT SOLD - ", "")
### Bearing
strwb_survey <- strwb_survey |> mutate(Bearing = temp2, .after = temp2)
strwb_survey$Bearing <- strwb_survey$Bearing |> 
                                str_replace( "^MEA.*", "") 
strwb_survey$Bearing[is.na(strwb_survey$Bearing)] <- ""
strwb_survey$temp2 <- strwb_survey$temp2 |> str_replace("^B.*", "")
strwb_survey$Bearing <- strwb_survey$Bearing |> str_replace("^BEARING - ", "")
#strwb_survey |> distinct(temp2)
## temp2 clear, except MEASURED IN 
## temp3
#strwb_survey |> distinct(temp3)
## temp3 contains "Utilized" and "Measured in xx" to separate
strwb_survey$temp3[is.na(strwb_survey$temp3)] <- ""  ## prepare for str_detect
strwb_survey$Utilized[str_detect(strwb_survey$temp3, "^U.*")] <- strwb_survey$temp3[str_detect(strwb_survey$temp3, "^U.*")]
strwb_survey$temp3 <- strwb_survey$temp3 |> str_replace("^U.*", "")
#strwb_survey |> distinct(temp3)
## temp3 clear, except MEASURED IN 
## combine temp2 and 3
strwb_survey <- strwb_survey |> unite(temp2, temp3, col="unit_desc", sep="")
## add "MEASURED IN xx" from temp4
strwb_survey$temp4[is.na(strwb_survey$temp4)] <- ""
strwb_survey$unit_desc[str_detect(strwb_survey$temp4, "^MEA.*")] <- strwb_survey$temp4[str_detect(strwb_survey$temp4, "^MEA.*")]
strwb_survey$temp4 <- strwb_survey$temp4 |> str_replace("^MEA.*", "")
strwb_survey$Utilized <- strwb_survey$Utilized |> str_replace("^UTILIZED - ", "")
strwb_survey$unit_desc <- strwb_survey$unit_desc |> str_replace("^MEASURED IN ", "")
#strwb_survey <- strwb_survey %>% select(-temp3)

```

<!-- Separate Value by units -->

```{r}
d <- data.frame(dcomma(strwb_survey$Value))
strwb_survey$Value <- d[,1]
strwb_survey$fn <- d[,2]

## Separate values by units($, CWT, TONS, LB, etc.)
strwb_survey$`Value($)` <- ifelse(strwb_survey$unit_desc == "$", strwb_survey$Value, "")
strwb_survey$`Value(CWT)` <- ifelse(strwb_survey$unit_desc == "CWT", strwb_survey$Value, "")
strwb_survey$`Value($/CWT)` <- ifelse(strwb_survey$unit_desc == "$ / CWT", strwb_survey$Value, "")
strwb_survey$`Value(TONS)` <- ifelse(strwb_survey$unit_desc == "TONS", strwb_survey$Value, "")
strwb_survey$`Value(LB)` <- ifelse(strwb_survey$unit_desc == "LB", strwb_survey$Value, "")
strwb_survey$`Value(LB/ACRE/APP)` <- ifelse(strwb_survey$unit_desc == "LB / ACRE / APPLICATION", strwb_survey$Value, "")
strwb_survey$`Value(LB/ACRE/YR)` <- ifelse(strwb_survey$unit_desc == "LB / ACRE / YEAR", strwb_survey$Value, "")
strwb_survey$`Value(NUMBER)` <- ifelse(strwb_survey$unit_desc == "NUMBER", strwb_survey$Value, "")
strwb_survey$`Value($/TON)` <- ifelse(strwb_survey$unit_desc == "$ / TON", strwb_survey$Value, "")
```

<!-- Domain -->

```{r}
strwb_survey <- strwb_survey |>
  separate_wider_delim(  cols = Domain,
                         delim = ",",
                         names = c("temp22",
                                 "temp23"),
                         too_many = "error",
                         too_few = "align_start"
                       )
#t22 <- unique(strwb_survey$temp22)
#t23 <- unique(strwb_survey$temp23)
strwb_survey$temp23[is.na(strwb_survey$temp23)] <- ""
strwb_survey$temp23 <- trimws(strwb_survey$temp23)
strwb_survey$temp22 <- trimws(strwb_survey$temp22)
```

<!-- Domain Category & PC code extraction -->

```{r}
strwb_survey <- strwb_survey |>
  separate_wider_delim(  cols = `Domain.Category`,
                         delim = ",",
                         names = c("temp42",
                                 "temp43",
                                 "temp44",
                                 "temp45"),
                         too_many = "error",
                         too_few = "align_start"
                       )

### Extract PC codes from temp43-45
strwb_survey$PC_t43 <- str_extract_all(strwb_survey$temp43, "= \\d+")
strwb_survey$PC_t44 <- str_extract_all(strwb_survey$temp44, "= \\d+")
strwb_survey$PC_t45 <- str_extract_all(strwb_survey$temp45, "= \\d+")
strwb_survey$PC_t43 <- gsub("= ", "", strwb_survey$PC_t43)
strwb_survey$PC_t44 <- gsub("= ", "", strwb_survey$PC_t44)
strwb_survey$PC_t45 <- gsub("= ", "", strwb_survey$PC_t45)
### substitute NAs
strwb_survey$PC_t43[is.na(strwb_survey$PC_t43)] <- ""
strwb_survey$PC_t44[is.na(strwb_survey$PC_t44)] <- ""
strwb_survey$PC_t45[is.na(strwb_survey$PC_t45)] <- ""
### combine t43-45 into PC code

strwb_survey <- strwb_survey |> unite(PC_t43, PC_t44, col="PC code", sep="")
strwb_survey <- strwb_survey |> unite(`PC code`, PC_t45, col="PC code", sep="")
strwb_survey$`PC code` <- gsub("character(0)", "", strwb_survey$`PC code`)

```

<!-- PC code into CAS number -->

```{r}
library(httr)
library(jsonlite)

cas <- function(c){
    path <- paste0("https://ordspub.epa.gov/ords/pesticides/apprilapi/?q=%7b%22ais%22:%7b%22$instr%22:%22",c,"%22%7d%7d")
    r <- GET(url = path)
    status_code(r)
    str(content(r))
    r_text <- content(r, as = "text", encoding = "UTF-8")
    df <- fromJSON(r_text,flatten = TRUE)
}

```


<!-- Separate survey into chem and market(mkt) -->

```{r}
strwb_survey_chem <- strwb_survey |> filter((temp22 == "CHEMICAL") | (temp42 == "CHEMICAL"))
strwb_survey_mkt <- strwb_survey |> filter(!((temp22 == "CHEMICAL") | (temp42 == "CHEMICAL")))
strwb_survey_mkt <- strwb_survey_mkt %>% select_if(~length(unique(.)) > 1)
strwb_survey_chem <- strwb_survey_chem %>% select_if(~length(unique(.)) > 1)
#strwb_survey |> distinct(temp45)
```





## References
